Stopping power of hafnium by proton impact and \\ the importance of relativistic $4f$ electrons

### 2020-04-11 response.tex

\documentclass[a4paper,10pt]{article}
\usepackage[utf8]{inputenc}
\setlength{\parindent}{0pt}
\usepackage{soul}
\usepackage{xcolor}

\setstcolor{red}

% Title Page
\title{Response to referee reports \\ AP11756/Montanari }
\author{C.C. Montanari, A. M. P. Mendez, D. M. Mitnik, J. E. Miraglia, \\
P.A. Miranda, R. Correa, J. Wachter, M. Aguilera, \\
E. Alves, N. Catarino and R.C. da Silva}
\date{\today}

\begin{document}
\maketitle

% ----------------------------------------------------------------------
\section{Reply to report of the First Referee}
% ----------------------------------------------------------------------

\textsl{1. Since only protons are considered as projectiles, the title 
needs to be modified accordingly.}

\vspace{0.1cm}
The title has been modified following the referee's recommendation to
``Stopping power of hafnium by proton impact and the importance of 
relativistic $4f$ electrons''.

% ----------------------------------------------------------------------
\vspace{0.25cm}
\textsl{2. It appears that calculations assume bare protons. According 
to CasP, neutrals play a considerable role in charge equilibrium around 
and below the Bragg peak. Some explanation is needed here.}

\vspace{0.1cm}
{\color{red}Reply here.}

% ----------------------------------------------------------------------
\vspace{0.25cm}
\textsl{3. Thickness measurements by Rutherford backscattering rely 
‘heavily’ on the scattering cross section, while energy loss is a minor 
correction depending on foil thickness (p.1, 2nd paragraph).}

\vspace{0.1cm}
{\color{red}Reply here.}

% ----------------------------------------------------------------------
\vspace{0.25cm}
\textsl{4. In order to get a more precise thickness value of the Hf 
foil, the authors use the energy loss of 5.486 MeV alphas as a standard, 
with SRIM as a reference. SRIM is based on empirical data, and for alpha 
particles in Hf, the IAEA database lists one single data point for He 
in Hf. Considering the spread in published stopping cross sections for 
e.g. He in Au, listed in the IAEA database, some explanation is needed 
to justify an error estimate of only 5\% (1st paragraph in section 
IIB).}

\vspace{0.1cm}
{\color{red}Reply here.}

% ----------------------------------------------------------------------
\vspace{0.25cm}
\textsl{5. In connection with the discussion of relativity and 
outer-shell electrons (p.3) it might be instructive to mention that 
this is the reason for the existence of relativistic quantum chemistry 
programs. Since the authors find it surprising that the effect increases 
in magnitude from inner to outer shells, the reader needs an 
explanation. I should think the effect of deviations in the screening 
add up shell by shell, but I may be wrong. Moreover, the sign of the 
discrepancy in figure 2a deserves attention, which seems to change from 
positive up to 4p- to negative from 5p+.}

\vspace{0.1cm}
The referee is correct. We included the following modifications in the text
(lines 200-214):

\vspace{0.1cm}
{\small ``One would expect the relativistic contributions to only affect 
significantly the inner shells. The relativistic binding energies of the 
electrons belonging to these shells are 2.5\% from the experimental values,
while the non-relativistic ones are within $\sim$10\%. Moreover, we notice 
the relativistic corrections greatly influences the outer $5p$ and $4f$ 
shells. This effect may be a consequence of the orthogonality with the 
relativistic inner shells. Also, the sign of the binding energy deviations 
is inverted near the FEG limit, i.e. the electrons go from being more 
bounded --than the experimental values-- to being less bounded. 
This change of sign can be attributed to the fact that the experimental 
values correspond to hafnium in solid-state, while our theoretical 
calculations correspond to the element in the gas phase.''}

% ----------------------------------------------------------------------
\vspace{0.25cm}
\textsl{6. The importance of 4f-5p screening is documented clearly 
(p.3), but the reader will ask why this does not affect other shells. 
I assume that the effect is important when two or more subshells have 
similar energies and, hence, overlap in real and velocity space. If so, 
I suggest to mention this. If not, another explanation is needed.}

\vspace{0.1cm}
{\color{red}Reply here.}

% ----------------------------------------------------------------------
\vspace{0.25cm}
\textsl{7. In view of the large contribution of the outermost shell, I 
suggest to include the core contribution in figure 3.}

\vspace{0.1cm}
{\color{red}Reply here.}

% ----------------------------------------------------------------------
\vspace{0.25cm}
\textsl{8. In addition to comparisons with experiments and empirical 
models, it would be instructive to see comparisons with theoretical 
models like CasP and DPASS in figure 4.}

\vspace{0.1cm}
{\color{red}Reply here.}

\newpage
% ----------------------------------------------------------------------
\section{Reply to report of the Second Referee}
% ----------------------------------------------------------------------

\textsl{1. Can the authors discuss how possible impurities in the sample and -especially- on both sides of the target would affect the final
stopping data? The authors must discuss it in the manuscript.}

\vspace{0.1cm}
{\color{red}Reply here.}

% ----------------------------------------------------------------------
\vspace{0.25cm}
\textsl{2. In the same context, the authors have to clarify whether potential
surface roughness and/or target non-uniformity might affect the final
results obtained in transmission method.}

\vspace{0.1cm}
{\color{red}Reply here.}

% ----------------------------------------------------------------------
\vspace{0.25cm}
\textsl{3. What is accuracy of the primary beam energy of the accelerator?
Moreover: when the target is placed and removed from the front of the
detector, does the detector itself need to unbiased and the chamber
opened? If so, might there be any offset ($\sim$keV) in the detection
system that would affect the energy distribution peaks, hence the
Gaussian position from the fits (Fig. 1)?}

\vspace{0.1cm}
{\color{red}Reply here.}

% ----------------------------------------------------------------------
\vspace{0.25cm}
\textsl{4. The experimental data above $\sim$1.5 MeV is systematically 
below the data from Ref. [7], hence SRIM semi-empirical approach, and 
also below the authors’ theory output. At these energies (and higher), 
the agreement between experiment and theory is expected to be better than
a few percent (that is why you used stopping power for 5.486 MeV alpha
on Hf to obtain the target thickness). Can the authors explain why
there is such systematic difference for the present measurements using
protons?}

\vspace{0.1cm}
{\color{red}Reply here.}

% ----------------------------------------------------------------------
\vspace{0.25cm}
\textsl{5. The authors claim an overall uncertainty of 5\%, but it is not clear
how they obtained this value, or whether only statistical
contributions were taken into account. Nevertheless, uncertainties of
measurements have to be treated following the Guide to the Expression
of Uncertainty in Measurement, International Organization for
Standardization, Geneva Switzerland, 1995. Especially for stopping
power measurements in transmission approach, there are other sources
of uncertainties that might potentially affect the final results
systematically (see comments 1, 2 and 3), and they must be discussed
in more details.}

\vspace{0.1cm}
{\color{red}Reply here.}

% ----------------------------------------------------------------------
\vspace{0.25cm}
\textsl{6. Can the present theory in the manuscript be compared to other
models (e.g., CasP and DPASS), especially towards low energies? Aiming
thus a constructive discussion on the overall-agreement among
different models to predict SCS of “challenge” materials, as
transition metals (f-subshell)?}

\vspace{0.1cm}
{\color{red}Reply here.}

% ----------------------------------------------------------------------
\vspace{0.25cm}
\textsl{7. In the last sentence of Sec. IV, the authors point out for a need
of more experimental data below 100 keV H+ but, in my opinion, more
data are needed even at higher energies, e.g., nearby (and slightly
above) the Bragg peak... Please enhance discussions about it.}

\vspace{0.1cm}
{\color{red}Reply here.}

\newpage
% ----------------------------------------------------------------------
\section{Reply to report of the Third Referee}
% ----------------------------------------------------------------------

\textsl{
- Regarding the theoretical model that describes the contribution to
the stopping power of the bound electrons, the authors emphasize the
importance of including relativistic effects in the description of the
binding energies and densities of the Hf 1s-4f states. However, it is
not clear if the calculations performed by the authors are done for
gas phase Hf atoms or for the solid Hf as in the experiments of Ref.
[38]. In case of being for the former, the slightly worse agreement
found for the 4f states as compared to the experimental values could
be related to this point.}

\vspace{0.1cm}
Our calculations are performed for gas phase Hf atoms. We have clarify
this in the manuscript (lines 211-214).

% ----------------------------------------------------------------------
\vspace{0.25cm}
\textsl{
1. The results from the three different theoretical calculations shown
in Fig. 3 are not completely clear. The authors emphasize the
importance of considering both screening effects and relativistic
effects for the 4f and 5p electrons, but it is not clear if later in
the text the authors use indistinctly the terminology ``relativistic''
and ``screening'' to refer to the same issue. For instance, are the ML
results shown in Fig. 3 calculated with the relativistic binding
energies and densities, but with and without including screening
corrections, or do the two ML curves differ in whether the
relativistic corrections are or not included? It is neither clear what
description of the bound electrons is used in the curve labeled as
``SPCC(FEG)+SLPA(bound)''.}

\vspace{0.1cm}
{\color{red}Reply here.}

% ----------------------------------------------------------------------
\vspace{0.25cm}
\textsl{
2. If all the results of Fig. 3 were calculated for the relativistic
description, it would be also interesting to show the results obtained
with the non-relativistic description in order to demonstrate the
importance of this correction.}

\vspace{0.1cm}
{\color{red}Te mando adjunto las funciones de onda y energías de ligaduras
del caso no relativista (NR). Habría que ver como queda.}

% ----------------------------------------------------------------------
\vspace{0.25cm}
\textsl{
3. Furthermore, it would also be meaningful to show the results from
the ML(FEG) calculation alone in order to identify better the
contribution of the bound electrons to the SP.}

\vspace{0.1cm}
{\color{red}Reply here.}

% ----------------------------------------------------------------------
\vspace{0.25cm}
\textsl{
4. When comparing the new and existing experimental data, clarify what
experimental technique was used in Ref. 7 to measure the stopping
power.}

\vspace{0.1cm}
{\color{red}Reply here.}

% ----------------------------------------------------------------------
\vspace{0.25cm}
\textsl{
5. The data by Sirotonin et al. plotted in Fig. 4 seem to suggest that
the stopping power maximum would be around 100-200 keV, while the
maximum predicted in the present work by the theoretical calculations
is located at slightly smaller energies. Please comment on this point
and discuss possible reasons for this discrepancy: What are the main
limitations in the theoretical model that can be causing this energy
shift?}

\vspace{0.1cm}
{\color{red}Reply here.}

% ----------------------------------------------------------------------
\vspace{0.25cm}
\textsl{
6. When reviewing on the experimental work performed on Hf oxides in
the introduction, note that there is a quite recent work by Roth and
co-workers published in Phys. Rev. Lett. 119, 163401 (2017), in which
the SP of protons is precisely measured in this target.}

\vspace{0.1cm}
We thank the referee for suggesting the missing reference. We have now
include it in the revised manuscript.

% ----------------------------------------------------------------------
\vspace{0.25cm}
\textsl{Finally, there are few misprints in the generated pdf file:}

\textsl{- Y-axis labels in Figs. 3 an 4 are unreadable }
{\color{red} ¿Cómo quedó el preprint? El pdf que genera mi latex tiene 
bien los labels, así que debe un problema del generador de latex del PRA...}

\st{\textsl{- Page 3,2nd column, 2nd full paragraph: "Too assess"}}

\st{\textsl{- Caption of Fig.2: "hollow circles" -$>$ "open circles" or
"unfilled circles"}}

\end{document}

### 2020-04-11 sp_hafnium.tex
include changes upon referees' comments

%%   This file is part of the APS files in the REVTeX 4 distribution.
%%   Version 4.1r of REVTeX, August 2010
%%
%%
%%   Copyright (c) 2001, 2009, 2010 The American Physical Society.
%%
%
\documentclass[aps,pra,reprint,groupedaddress]{revtex4-1}

%\documentclass[aps,pra,twocolumn,groupedaddress]{revtex4-1}
%\documentclass[aps,prb,reprint,superscriptaddress]{revtex4-1}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{amsmath} 
\usepackage{graphicx}
\usepackage[mathlines]{lineno}% Enable numbering of text and display math
%\linenumbers\relax % Commence numbering lines

@@

%Title of paper
\title{The stopping power of hafnium and the importance of relativistic $4f$ electrons}
%Authors and affiliations

%---------------------------Argentina
\author{C. C. Montanari}
\email{mclaudia@iafe.uba.ar}
\affiliation{Instituto de
Astronom\'{\i}a y F\'{\i}sica del Espacio, Consejo Nacional de Investigaciones Cient\'{\i}ficas y
T\'{e}cnicas - Universidad de Buenos Aires, Pabell\'on IAFE, 1428 Buenos Aires, Argentina}%
%\affiliation{Universidad de Tres de Febrero, Sáenz Peña, Buenos Aires, Argentina.}
%------------------------------
\author{A. M. P. Mendez}
\affiliation{Instituto de
Astronom\'{\i}a y F\'{\i}sica del Espacio, Consejo Nacional de Investigaciones Cient\'{\i}ficas y
T\'{e}cnicas - Universidad de Buenos Aires, Pabell\'on IAFE, 1428 Buenos Aires, Argentina}%
%-------------------------------
\author{D. M. Mitnik}
\affiliation{Instituto de
Astronom\'{\i}a y F\'{\i}sica del Espacio, Consejo Nacional de Investigaciones Cient\'{\i}ficas y
T\'{e}cnicas - Universidad de Buenos Aires, Pabell\'on IAFE, 1428 Buenos Aires, Argentina}%
%\affiliation{Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales, Departamento de F\'{\i}sica, Ciudad Universitaria, 1428 Buenos Aires, Argentina.}
%----------------------------------------------
\author{J. E. Miraglia}
\affiliation{Instituto de
Astronom\'{\i}a y F\'{\i}sica del Espacio, Consejo Nacional de Investigaciones Cient\'{\i}ficas y
T\'{e}cnicas - Universidad de Buenos Aires, Pabell\'on IAFE, 1428 Buenos Aires, Argentina}%
%---------------------------Chile

\author{P.A. Miranda}
\affiliation{Departmento de F\'isica, Facultad de Ciencias Naturales, Matem\'atica y del Medio Ambiente. Universidad Tecnol\'ogica Metropolitana. 7800002, Chile}
%----------------------------
\author{R. Correa}
\affiliation{Departmento de F\'isica, Facultad de Ciencias Naturales, Matem\'atica y del Medio Ambiente. Universidad Tecnol\'ogica Metropolitana. 7800002, Chile}
\author{J. Wachter}
\affiliation{Departmento de F\'isica, Facultad de Ciencias Naturales, Matem\'atica y del Medio Ambiente. Universidad Tecnol\'ogica Metropolitana. 7800002, Chile}
\author{M. Aguilera}
\affiliation{Departmento de F\'isica, Facultad de Ciencias Naturales, Matem\'atica y del Medio Ambiente. Universidad Tecnol\'ogica Metropolitana. 7800002, Chile}
%-------------------------Lisboa
\author{E. Alves}
\affiliation{Centro de Ci\^{e}ncias e Tecnologias Nucleares, Instituto Superior T\'ecnico, Universidade de Lisboa,
2696-953 Sacav\'{e}m, Portugal}
\affiliation{Instituto de Plasma e Fus\~{a}o Nuclear, Instituto Superior T\'ecnico, Universidade de Lisboa, 2696-953 Sacav\'{e}m, Portugal}
%---------------------------
\author{N. Catarino}
\affiliation{Centro de Ci\^{e}ncias e Tecnologias Nucleares, Instituto Superior T\'ecnico, Universidade de Lisboa,
2696-953 Sacav\'{e}m, Portugal}
%---------------------------
\author{R.C. da Silva}
\affiliation{Centro de Ci\^{e}ncias e Tecnologias Nucleares, Instituto Superior T\'ecnico, Universidade de Lisboa,
2696-953 Sacav\'{e}m, Portugal}

@@

%=============================================================================================================================================================================================
%=============================================================================================================================================================================================
\begin{abstract}
 The stopping power of protons through Hf foil has been studied both experimentally and theoretically. The measurements were performed at the Laboratory of Accelerators and X-Ray Diffraction in Lisbon, by using the transmission method on self-supporting stopping material. %for (0.6-2.5) MeV protons. 
 The overall uncertainty of around $5\%$ was established over the protons energy range  (0.6-2.5) MeV. The theoretical developments involved fully relativistic atomic structure calculations for %this Z=72 target 
 Hf, which required the solution of the Dirac equation. The shell-wise local plasma approximation (SLPA) was used to describe the energy transferred to the bound $1s$-$4f$ electrons, and the outer four electrons were considered as a free electron gas (FEG). %We contemplated two different models for the FEG: the screened potential with cusp condition (SPCC) for energies below that of the plasmon excitation, and the dielectric formalism, for energies around the stopping maximum and above. 
 We found the relativistic description of the $4f$-shell and the screening between $4f$ and $5p$ electrons are decisive around the stopping maximum. %for energies above 40 keV.
 Present theoretical and experimental results are in very good agreement in the energy region of the new measurements. However, the present theoretical stopping cross sections show %interesting 
 substantial differences with the most used semi-empirical models (SRIM2013 and ICRU-49) at intermediate to low energies. Our calculations suggest the stopping maximum %around 40 eV$/(10^{15}$ atom/cm$^2)$ at 70 keV, which is
 higher and shifted to lower energies than these previous predictions. Future measurements below 100 keV would be decisive for a better knowledge  around the maximum and below.
\end{abstract}

@@

%\linenumbers
%=============================================================================================================================================================================================
%=============================================================================================================================================================================================
\section{Introduction}
\label{intro}

For impact energies above a few keV/amu, mono-energetic charged particles penetrating a foil of any material lose their energy through a series of consecutive inelastic collisions, mainly with target electrons \cite{Chu01,Sigmund}. The information given by the energy loss process is essential not only to have a better knowledge of the physics behind the fundamental interactions, but also because it plays an important role in many applied fields such as materials science, nuclear physics, ionic implantation and radiotherapy \cite{Sigmund,Schardt}. Experimental data on ion mean energy loss per unit path $S(E)$ is of crucial relevance to check the reliability of semi-empirical models and to determine some key parameters 
%included in theoretical approaches of stopping power %\cite{Diwan,Damache01,Damache02,Damache03,Damache04}. 
\cite{Diwan,Damache04,Damache02}. The experimental data available is often rather scarce, which is %particularly 
troublesome when the material under study corresponds to an element of low occurrence on the Earth's upper crust, such as hafnium.

So far, only one experimental work has been published regarding the stopping power cross section of pure hafnium for protons \cite{Sirotinin}, while more attention has been recently given to studies involving hafnium oxide due to its practical use \cite{Abril,Behar,Primetzhofer}. It is well known that significant attention has been paid in recent years to transition metal-oxides such as HfO$_2$ because of their potential as alternative gate dielectrics to replace SiO$_2$ for the future generation of nano-electronics with less than 45 nm gate length \cite{Choi,Robertson}.  Some important physical properties of the above mentioned metal-oxide films depend on its thickness, which is often measured by using Rutherford Backscattering Spectrometry \cite{Alfassi01, Tesmer01}, a method that relies heavily on the determination of the stopping power of ion beams in the material of interest.

In this study, we report experimental stopping power cross sections over the incident energy range (0.6-2.5) MeV for protons crossing self-supported Hf thin-film by using the transmission method. We aim not only to upgrade stopping power data compilations \cite{HPaul03,mondim17} but also to provide useful information about the processes governing the slowing down of protons in multi-electronic targets. In the rare earth metals, the $4f$ electrons play a relevant role in the stopping power, being the first shell of bound electrons below the conduction band. As already noted \cite{Roth17}, the free electron gas (FEG) shows unexpected %a peculiar
behavior in these elements, which casts %that states 
doubts on its proper description. In the case of Hf, we found the contribution of the $4f$-shell decisive even at impact energies around the stopping maximum, as shown later.

The theoretical approach implemented in this work uses the shell-wise local plasma approximation (SLPA) \cite{mon13} to describe the energy transferred to the bound $1s$-$4f$ electrons, and two different models for the FEG: the screened potential with cusp condition model (SPCC) \cite{mon17}, which is non-linear binary formalism%,  for energies below that of the plasmon excitation
; and the Mermin-Lindhard dielectric formalism (ML) \cite{Mermin}, for energies around the stopping maximum and above. Our model requires 
the relativistic wave functions and binding energies of Hf, and considers 4 electrons per atom in the  FEG \cite{mendez2019}. 
%On the other hand,
Hf has the extra interest of the filled $4f$-subshell (with 14 electrons) as the main contribution below the FEG, causing the stopping cross sections to be very sensitive to a good description of this shell. The screening among the $4f$ and $5p$ electrons has been considered and found to play a major role within the SLPA calculations.

The experimental details and data are given in Section \ref{experiment}, while the theoretical method is explained in Section \ref{theory}. Present theoretical and experimental values are finally compared to the available experimental values \cite{Sirotinin},  the ICRU-49 calculations \cite{ICRU49} and the SRIM-2013 package \cite{Ziegler01}. Conclusions and discussions are given in Section \ref{conclusion}. All the present data can be found at the Zenodo platform \cite{zenodo}.

%=============================================================================================================================================================================================

\begin{figure}[!t]
\centering
\includegraphics[width=9cm]{Fig01.eps}
\caption{RBS spectrum for $E_{\mathrm{avg}} = 921.1$ keV protons on hafnium sample which is subsequently used to determine the energy loss in the foil.}
\label{F01}
\end{figure}

%=============================================================================================================================================================================================

\section{EXPERIMENTAL ARRANGEMENTS}
\label{experiment}
\subsection{Accelerator and scattering Chamber}
The procedure used in this work to obtain stopping power data is essentially the same as described in Ref. \cite{Miranda01}. The present measurements were made at the IST/LATR
(Laboratory of Accelerators and X-Ray Diffraction) in Lisbon. This facility uses a 2.5 MV Van de Graaff accelerator to deliver $^1$H$^+$ primary ion beams through a series of electrostatic lenses and collimators onto a thin Au/SiO$_2$ sample which is used as a scattering center. This sample was placed in the center of a RBS/C scattering chamber, where a high vacuum (pressure of
$\sim$10$^{-6}$ Torr) was maintained during the measurements. The beam current on the sample was kept at around 5.0 nA to attain sufficient statistics in each particle spectrum.
By using a beam spot of about 1.0 mm in diameter, a solid angle of 11.4 msr was attained. The overall energy resolution (FWHM) of the detection system was about 15 keV relative to 5.486 MeV alpha particles from a $^{241}$Am source.

>>>>

\documentclass[aps,pra,reprint,groupedaddress,showpacs,showkeys]{revtex4-1}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{amsmath} 
\usepackage{graphicx}
\usepackage{soul}
\usepackage{xcolor}
\setstcolor{red}
\usepackage[mathlines]{lineno}% Enable numbering of text and display math
%\linenumbers\relax % Commence numbering lines

@@

%Title of paper
\title{Stopping power of hafnium by proton impact and \\
the importance of relativistic $4f$ electrons}
%Authors and affiliations

%---------------------------Argentina
\author{C. C. Montanari}
\email{mclaudia@iafe.uba.ar}
\affiliation{Instituto de Astronom\'{\i}a y F\'{\i}sica del Espacio, 
Consejo Nacional de Investigaciones Cient\'{\i}ficas y T\'{e}cnicas - 
Universidad de Buenos Aires, Pabell\'on IAFE, 1428 Buenos Aires, 
Argentina}
%------------------------------
\author{A. M. P. Mendez}
\affiliation{Instituto de Astronom\'{\i}a y F\'{\i}sica del Espacio, 
Consejo Nacional de Investigaciones Cient\'{\i}ficas y T\'{e}cnicas - 
Universidad de Buenos Aires, Pabell\'on IAFE, 1428 Buenos Aires, 
Argentina}
%-------------------------------
\author{D. M. Mitnik}
\affiliation{Instituto de Astronom\'{\i}a y F\'{\i}sica del Espacio, 
Consejo Nacional de Investigaciones Cient\'{\i}ficas y T\'{e}cnicas - 
Universidad de Buenos Aires, Pabell\'on IAFE, 1428 Buenos Aires, 
Argentina}
%----------------------------------------------
\author{J. E. Miraglia}
\affiliation{Instituto de Astronom\'{\i}a y F\'{\i}sica del Espacio, 
Consejo Nacional de Investigaciones Cient\'{\i}ficas y T\'{e}cnicas - 
Universidad de Buenos Aires, Pabell\'on IAFE, 1428 Buenos Aires, 
Argentina}
%---------------------------Chile

\author{P.A. Miranda}
\affiliation{Departamento de F\'isica, Facultad de Ciencias Naturales, 
Matem\'atica y del Medio Ambiente. Universidad Tecnol\'ogica Metropolitana. 
7800002, Chile}
%----------------------------
\author{R. Correa}
\affiliation{Departamento de F\'isica, Facultad de Ciencias Naturales, 
Matem\'atica y del Medio Ambiente. Universidad Tecnol\'ogica Metropolitana. 
7800002, Chile}
%----------------------------
\author{J. Wachter}
\affiliation{Departamento de F\'isica, Facultad de Ciencias Naturales, 
Matem\'atica y del Medio Ambiente. Universidad Tecnol\'ogica Metropolitana. 
7800002, Chile}
%----------------------------
\author{M. Aguilera}
\affiliation{Departamento de F\'isica, Facultad de Ciencias Naturales, 
Matem\'atica y del Medio Ambiente. Universidad Tecnol\'ogica Metropolitana. 
7800002, Chile}
%-------------------------Lisboa
\author{E. Alves}
\affiliation{Centro de Ci\^{e}ncias e Tecnologias Nucleares, Instituto 
Superior T\'ecnico, Universidade de Lisboa, 2696-953 Sacav\'{e}m, 
Portugal}
\affiliation{Instituto de Plasma e Fus\~{a}o Nuclear, Instituto Superior 
T\'ecnico, Universidade de Lisboa, 2696-953 Sacav\'{e}m, Portugal}
%---------------------------
\author{N. Catarino}
\affiliation{Centro de Ci\^{e}ncias e Tecnologias Nucleares, Instituto 
Superior T\'ecnico, Universidade de Lisboa, 2696-953 Sacav\'{e}m, 
Portugal}
%---------------------------
\author{R.C. da Silva}
\affiliation{Centro de Ci\^{e}ncias e Tecnologias Nucleares, Instituto 
Superior T\'ecnico, Universidade de Lisboa, 2696-953 Sacav\'{e}m, 
Portugal}

@@

%=======================================================================
%=======================================================================
\begin{abstract}
 The stopping power of protons through Hf foil has been studied both 
experimentally and theoretically. The measurements were performed at 
the Laboratory of Accelerators and X-Ray Diffraction in Lisbon, by 
using the transmission method on self-supporting stopping material. 
 The overall uncertainty of around $5\%$ was established over the 
protons energy range  (0.6-2.5) MeV. The theoretical developments 
involved fully relativistic atomic structure calculations for Hf, which 
required the solution of the Dirac equation. The shell-wise local plasma 
approximation (SLPA) was used to describe the energy transferred to the 
bound $1s$-$4f$ electrons, and the outer four electrons were considered 
as a free electron gas (FEG). We found the relativistic description of 
the $4f$-shell and the screening between $4f$ and $5p$ electrons are 
decisive around the stopping maximum. Present theoretical and 
experimental results are in very good agreement in the energy region 
of the new measurements. However, the present theoretical stopping 
cross sections show substantial differences with the most used 
semi-empirical models (SRIM2013 and ICRU-49) at intermediate to low 
energies. Our calculations suggest the stopping maximum higher and 
shifted to lower energies than these previous predictions. Future 
measurements below 100 keV would be decisive for a better knowledge 
around the maximum and below.
\end{abstract}

@@

\linenumbers
%=======================================================================
%=======================================================================
\section{Introduction}
\label{intro}

For impact energies above a few keV/amu, mono-energetic charged 
particles penetrating a foil of any material lose their energy through 
a series of consecutive inelastic collisions, mainly with target 
electrons~\cite{Chu01,Sigmund}. The information given by the energy 
loss process is essential not only to have a better knowledge of the 
physics behind the fundamental interactions, but also because it plays 
an important role in many applied fields such as materials science, 
nuclear physics, ionic implantation and radiotherapy~\cite{Sigmund,Schardt}. 
Experimental data on ion mean energy loss per unit path $S(E)$ is of 
crucial relevance to check the reliability of semi-empirical models and 
to determine some key parameters~\cite{Diwan,Damache04,Damache02}. The 
experimental data available is often rather scarce, which is troublesome 
when the material under study corresponds to an element of low 
occurrence on the Earth's upper crust, such as hafnium.

So far, only one experimental work has been published regarding the 
stopping power cross section of pure hafnium for protons~\cite{Sirotinin}, 
while more attention has been recently given to studies involving 
hafnium oxide due to its practical use~\cite{Abril,Behar,Primetzhofer,Roth}. 
It is well known that significant attention has been paid in recent 
years to transition metal-oxides such as HfO$_2$ because of their 
potential as alternative gate dielectrics to replace SiO$_2$ for the 
future generation of nano-electronics with less than 45 nm gate 
length~\cite{Choi,Robertson}. Some important physical properties of the 
above mentioned metal-oxide films depend on its thickness, which is 
often measured by using Rutherford Backscattering Spectrometry~\cite{Alfassi01,Tesmer01}, 
a method that relies heavily on the determination of the stopping power 
of ion beams in the material of interest.

In this study, we report experimental stopping power cross sections 
over the incident energy range (0.6-2.5) MeV for protons crossing 
self-supported Hf thin-film by using the transmission method. We aim 
not only to upgrade stopping power data compilations~\cite{HPaul03,mondim17} 
but also to provide useful information about the processes governing 
the slowing down of protons in multi-electronic targets. In the rare 
earth metals, the $4f$ electrons play a relevant role in the stopping 
power, being the first shell of bound electrons below the conduction 
band. As already noted \cite{Roth17}, the free electron gas (FEG) shows 
unexpected behavior in these elements, which casts doubts on its proper 
description. In the case of Hf, we found the contribution of the 
$4f$-shell decisive even at impact energies around the stopping maximum, 
as shown later.

The theoretical approach implemented in this work uses the shell-wise 
local plasma approximation (SLPA)~\cite{mon13} to describe the energy 
transferred to the bound $1s$-$4f$ electrons, and two different models 
for the FEG: the screened potential with cusp condition model (SPCC)~\cite{mon17}, 
which is non-linear binary formalism; and the Mermin-Lindhard dielectric 
formalism (ML)~\cite{Mermin}, for energies around the stopping maximum 
and above. Our model requires the relativistic wave functions and 
binding energies of Hf, and considers 4 electrons per atom in the 
FEG~\cite{mendez2019}. Hf has the extra interest of the filled 
$4f$-subshell (with 14 electrons) as the main contribution below the 
FEG, causing the stopping cross sections to be very sensitive to a good 
description of this shell. The screening among the $4f$ and $5p$ 
electrons has been considered and found to play a major role within the 
SLPA calculations.

The experimental details and data are given in Section~\ref{experiment}, 
while the theoretical method is explained in Section~\ref{theory}. 
Present theoretical and experimental values are finally compared to the 
available experimental values~\cite{Sirotinin}, the ICRU-49 
calculations~\cite{ICRU49} and the SRIM-2013 package~\cite{Ziegler01}. 
Conclusions and discussions are given in Section~\ref{conclusion}. All 
the present data can be found at the Zenodo platform \cite{zenodo}.

%-----------------------------------------------------------------------
\begin{figure}[!t]
\centering
\includegraphics[width=9cm]{Fig01.eps}
\caption{RBS spectrum for $E_{\mathrm{avg}}=921.1$ keV protons on 
hafnium sample which is subsequently used to determine the energy loss 
in the foil.}
\label{F01}
\end{figure}

%=======================================================================
\section{EXPERIMENTAL ARRANGEMENTS}
\label{experiment}

%-----------------------------------------------------------------------
\subsection{Accelerator and scattering Chamber}
The procedure used in this work to obtain stopping power data is 
essentially the same as described in Ref.~\cite{Miranda01}. The present 
measurements were made at the IST/LATR (Laboratory of Accelerators and 
X-Ray Diffraction) in Lisbon. This facility uses a 2.5 MV Van de Graaff 
accelerator to deliver $^1$H$^+$ primary ion beams through a series of 
electrostatic lenses and collimators onto a thin Au/SiO$_2$ sample which 
is used as a scattering center. This sample was placed in the center of 
a RBS/C scattering chamber, where a high vacuum (pressure of 
$\sim$10$^{-6}$ Torr) was maintained during the measurements. The beam 
current on the sample was kept at around 5.0 nA to attain sufficient 
statistics in each particle spectrum. By using a beam spot of about 
1.0 mm in diameter, a solid angle of 11.4 msr was attained. The overall 
energy resolution (FWHM) of the detection system was about 15 keV 
relative to 5.486 MeV alpha particles from a $^{241}$Am source.

%-----------------------------------------------------------------------
\subsection{Target}
The stopping material under analysis was a hafnium foil with nominal 
thickness of 1.0 $\mu$m and 99.95\% purity which was supplied by Lebow 
Company~\cite{Lebow}. However, a more precise thickness value was 
achieved by measuring the energy loss of alpha particles coming from a 
calibrated ($^{239}$Pu, $^{241}$Am, $^{244}$Cm) source. From the alpha 
spectra with and without the Hf foil interposed, the characteristic 
energy shift $\delta$E was measured and then combined with the stopping 
power for 5.486 MeV alphas on hafnium (55.69 eV/$10^{15}$ at/cm$^2$) 
found in Ref.~\cite{Ziegler01} to obtain an areal density of 
($4.13 \pm 0.21$)$\times 10^{19}$ at/cm$^2$ which corresponds to a 
thickness of $0.920\pm0.046 \mu$m.

%-----------------------------------------------------------------------
\subsection{Energy loss measurement}
Once the beam impinges on the Au/SiO$_2$ sample, protons are 
back-scattered towards a Si surface barrier detector located at 
140$^{\circ}$ relative to the initial beam direction. Fig.~\ref{F01} 
shows two particle spectra where ion energies $E_2$ and $E_1$ are 
associated to a placed and removed hafnium sample, respectively. Both 
energy distributions were fitted by Gaussian functions to obtain the 
mean energy and width (FWHM) of the peaks \cite{Sun01}, and from the 
difference between these two peak positions in the spectrum, the total 
energy loss $\Delta E = (E_1 - E_2)$ in the foil was calculated. As 
established in previous studies \cite{Miranda01,Damache02}, the 
experimental stopping power cross sections $\varepsilon (E) $ are 
determined at some mean energy $E_{\mathrm{avg}}$ by measuring the ion 
energy losses $\Delta E$ within the investigated Hf foil, which has a 
mean thickness denoted by $\Delta x$. In this way, only when the energy 
loss fraction $\Delta E/E_{\mathrm{avg}}$ across the Hf foil is not 
exceeding 20\% it is possible to define the stopping cross section by 
\cite{Raisanen01,Schulz01}:
\begin{equation}\label{eq:stcross}
 \varepsilon(E)=\frac{S(E)}{N}=-\frac{dE}{N\,dx}\approx-\frac{\Delta E}{N\Delta x},
\end{equation}
where $N$ denotes the atomic number density (atoms cm$^{-3}$) of the 
material under study. When this condition was not fulfilled, a small 
correction to the mean energies $E_{\mathrm{avg}}$ was applied in order 
to account for the nonlinear dependence on ion energy of stopping 
powers \cite{Chilton,Rajatora}.
